---
title: 'Analyzing the affect of Redlining on LAs reported bird biodiversity'
author: "{Hazel Vaquero}"
output: html_document
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```


# Background Information

Redlining is a historical practice with racial undertones incorporated
in 1930 by the Home Owners' Loan Corporation (HOLC), as part of the New
Deal. Neighborhoods were rated based on their perceived safety for real
estate investment. Their ranking system, A (green), B (blue), C
(yellow), D (red)) was then used to block access to loans for home
ownership. This practice targeted and discriminated against minority
residents. This unjust practice halted economic development and created
environmental injustice[^1].

A recent study found that redlining is also affecting the observations
of biodiversity[^1]. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most under sampled areas across 195 US cities.
Observation like these rely on community or citizen scientist to
contribute and identify species. The gap in data is concerning because
conservation decisions are made based on these data.

[^1]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Specifically in LA county, neighborhoods in redlining district view less
biodiversity in birds[^2]. These areas have fewer canopy covers which
leads to only certain birds more resilient to urban life to be present.

[^2]: Eric M Wood, Sevan Esaian, Christian Benitez, Philip J Ethington,
    Travis Longcore, Lars Y Pomara, Historical racial redlining and
    contemporary patterns of income inequality negatively affect birds,
    their habitat, and people in Los Angeles, California, Ornithological
    Applications, 2023;, duad044,
    <https://doi.org/10.1093/ornithapp/duad044>

# Data

[EJScreen]{.underline}

The Environmental Protection Agency's Environmental Justice Screening and
Mapping Tool spatial data was used to access data on income, health, and race for 
LA county. The EJScreen uses data that is publicly available to be used to 
analyze environmental justice across the country. There are potential limitations
to this dataset: uncertainty in estimates for smaller areas, does not cover all 
relevant issues, and environmental indicators are only screening-level proxies
for actual health impacts.

EJScreen provides data on environmental and demographic information for
the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract)
and [block group](https://en.wikipedia.org/wiki/Census_block_group)
levels.

[Mapping Inequality]{.underline}

The redlining maps were obtain from the [Digital Scholarship
Lab](https://dsl.richmond.edu/). Researchers at the University of
Richmond have digitized maps and information from the HOLC as part of
the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project. It was directly accessed from the site and downloaded as a
ShapeFile.

Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads)[^3].

[^3]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

[Global Biodiversity Information Facility]{.underline}

The community scientist bird observations data was downloaded as a
ShapeFile from the [Global Biodiversity Information Facility](gbif.org).
Observations include the location, date, and species observed..

# Analysis:

My initial analysis is to get a scope of LA county economic and racial
distribution. Using the EJScreen I am interested in analyzing Census
block groups that:

    -   are above the 95th percentile of national values for wastewater
        discharge

    -   have less than 5% of the population are considered low-income

    -   above the 80th percentile for Particular Matters 2.5

    -   above the 80th percentile for Superfund proximity

Building on to these I will also analyze the economic and health
disparity in historical redlined areas. How does redlining areas relate
to:

-   rates of low income

-   percentile for Particular Matter 2.5

-   percentile for low life expectancy

-   percentile for air toxic cancer risk

```{r, message=FALSE, warning=FALSE}
library(tmap)
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)
library(gt)
library(magrittr)
library(formattable)
library(janitor)
```

```{r, warning = FALSE, error=FALSE, message=FALSE}
# Import EJ Screen data
ej_screen <-
  st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb") %>% 
  clean_names()

# Import redlining information for Los Angeles

LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>% 
  st_make_valid() # validates geometry
```

### Prepare data

Let's prepare the data for Los Angeles county and Redlining for
analysis.

```{r, message=FALSE, warning=FALSE, error=FALSE}
# Filter to Los Angeles and remove NA values
los_angeles <- ej_screen %>% filter(cnty_name == "Los Angeles County",
                                    !is.na(p_pwdis))

# Update CRS of Los Angeles and Redlining
if (st_crs(los_angeles) != st_crs(LA_redlining)){
  LA_redlining <- st_transform(LA_redlining, crs = st_crs(los_angeles))
}

# Check if CRS are equal
st_crs(los_angeles) == st_crs(LA_redlining)
```


```{r, include=FALSE}
# Filter census block groups above the 95th percentile of national values for 
# wastewater discharge
los_angeles_water <- los_angeles %>% 
  filter(p_pwdis > 95) %>% 
  st_centroid() # turn into a point polygon
```

Wastewater discharge
```{r, echo = FALSE}
# Los Angles centroid base
tm_shape(los_angeles) +
  tm_polygons(fill = 'p_pwdis',
              title = "Percentile of Wastewater Discharge") +
  tm_shape(los_angeles_water) + #wastewater discharge centroid
  tm_dots(
    'p_pwdis',
    alpha = 0.7,
    title.size = "A",
    col = 'red',
    title = 'A'
  ) +
  tm_layout(
    legend.outside = FALSE,
    main.title = 'LA Wastewater discharge above 95th percentile',
    main.title.size = 10,
    legend.text.size = 0.7,
    legend.title.size = 0.7
  ) +
  tm_compass(type = "8star",
             size = 1.5,
             position = c("left", "bottom")) +
  tm_graticules() +
  tm_scale_bar(position = c(0.7, 0)) +
  tm_add_legend(labels = 'Wastewater Discharge',
                fill = 'red')
```

How many census block groups have less than 5% of the population
considered low income?

```{r}
# filter census blocks
low_income_5 <- los_angeles %>%
  select(lowincpct) %>%
  filter(lowincpct < .05) 


# Percent of the population is low income
paste(round(nrow(low_income_5)/nrow(los_angeles) * 100,3), "% of the population is low income")
```

80th percentile for PM 2.5

```{r}
# above the 80th percentile for Particulate Matter 2.5 and above the 80th percentile for Superfund proximity
pm2_5 <- los_angeles %>% select(p_pm25, p_pnpl) %>%
  filter(p_pm25 > 80 & p_pnpl > 80) 

# Matter 2.5 and above the 80th percentile for Superfund proximity
paste(round(nrow(pm2_5)/nrow(los_angeles)*100,3), "% of the census block are above the 80th percentile for Particular Matter 2.5")
```

Make a map of historical redlining boundaries, colored by HOLC grade.

```{r}
tm_shape(LA_redlining) +
  tm_polygons("holc_grade",
              title = "HOLC grade") +
  tm_compass(type = "8star",
             size = 2) +
  tm_graticules() +
  tm_scale_bar() + 
  tm_layout(main.title = "LA redlining boundaries")
```
