---
title: 'Analyzing the affect of Redlining on LAs reported bird biodiversity'
author: "{Hazel Vaquero}"
output: html_document
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

# Background information

Current environmental injustice reflects legacies of the past. In the
1930's the Home Owners' Loan Corporation (HOLC), as part of the New
Deal, rated neighborhoods based on their perceived safety for real
estate investment. Their ranking system, (A (green), B (blue), C
(yellow), D (red)) was then used to block access to loans for home
ownership. This practice targeted and discriminated against minority
residents also known as redlining. This unjust practice is still felt to
this day as it halted economic development and impacted the health of
the residents[^1].

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

A recent study found that redlining is also affecting the observations
of biodiversity[^2]. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most under sampled areas across 195 US cities.
Observation like these rely on community or citizen scientist to
contribute and identify species. The gap in data is concerning because
conservation decisions are made based on these data.

[^2]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

A recent study has found that LA neighborhoods that are in a redlining
district view less diversity in birds[^3]. There are fewer canopy covers
in lower-income neighborhoods and the birds found are those that are
resilient to living in urban settings. Reducing the biodiversity of
birds that can survive in these neighborhoods.

[^3]: Eric M Wood, Sevan Esaian, Christian Benitez, Philip J Ethington,
    Travis Longcore, Lars Y Pomara, Historical racial redlining and
    contemporary patterns of income inequality negatively affect birds,
    their habitat, and people in Los Angeles, California, Ornithological
    Applications, 2023;, duad044,
    <https://doi.org/10.1093/ornithapp/duad044>

This notebook will conduct an analysis of environmental justice in L.A.
County's census block. As well as the relationship of historical
redlining districts to environmental injustices: pollution, income
disparities, and health effects. This will expand to an analysis of
historical redline communities(D) that are seeing lower reports of birds
than their counterparts (A).

# Data Used

1.  **EJScreen:** Environmental Protection Agency's Environmental
    Justice Screening and Mapping Tool. According to the [US EPA
    website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

    > EJScreen is an EPA's environmental justice mapping and screening
    > tool that provides EPA with a nationally consistent dataset and
    > approach for combining environmental and demographic socioeconomic
    > indicators. EJScreen users choose a geographic area; the tool then
    > provides demographic socioeconomic and environmental information
    > for that area. All of the EJScreen indicators are publicly
    > available data. EJScreen simply provides a way to display this
    > information and includes a method for combining environmental and
    > demographic indicators into EJ indexes.

EJScreen provides data on environmental and demographic information for
the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract)
and [block group](https://en.wikipedia.org/wiki/Census_block_group)
levels. This notebook will analyse:

-   Census block groups that

    -   are above the 95th percentile of national values for wastewater
        discharge

    -   have less than 5% of the population are considered low-income

    -   above the 80th percentile for Particular Matters 2.5

    -   above the 80th percentile for Superfund proximity

2.  **Mapping Inequality**

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads)[^4].

[^4]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

This notebook will cover:

Historical redlining areas

-   low income

-   percentile for Particular Matter 2.5

-   percentile for low life expectancy

-   percentile for air toxic cancer risk

3.  **Global Biodiversity Information Facility**

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

# Analysis

### Import Libraries
Load relevant packages

```{r, error=FALSE, warning=FALSE}
library(tmap)
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)
library(gt)
library(magrittr)
library(formattable)
library(janitor)
```

### EJScreen data analysis
Let's first read in the data and filter to Los Angeles County
```{r}
ej_screen <-
  st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb") %>% 
  clean_names()

# Filter to Los Angeles
los_angeles <- ej_screen %>% filter(cnty_name == "Los Angeles County")

```

