---
title: 'EDS 223: assignment 2'
author: "{Hazel Vaquero}"
output: html_document
date: "2023-10-17"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

#### EJScreen

We will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
You will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To
understand the associated data columns, you will need to explore the
Technical Documentation and column description spreadsheet available in
the data folder. I also encourage you to explore the [limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen)
of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.

## Assignment

### Investigate the legacy of redlining in current environmental (in)justice

Load relevant packages.

```{r }
library(tmap)
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)
library(gt)
library(magrittr)
library(formattable)
```

Read in EJScreen data and filter to Los Angeles County (5 points)

```{r }
# read in EJScreen data
ej_screen <- st_read("~/eds223/assignment-2-hazelvaq/data (1)/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")

# Filter to Los Angeles County
los_angeles <- ej_screen %>% filter(CNTY_NAME == "Los Angeles County")

```

Make a map of wastewater discharge by census block groups. Indicate
which census block groups are above the 95th percentile of national
values for wastewater discharge by adding a centroid. (10 points)

```{r }
## Filtering Los Angeles dataframe to only have rows that have values
# remove islands 
la <- los_angeles %>% 
  filter(!is.na(P_PWDIS)) 

# Filter census block groups above the 95th percentile of national values for 
# wastewater discharge
los_angeles_water <- los_angeles %>% 
  filter(P_PWDIS > 95) %>% 
  st_centroid()

# Map of Wastewater discharge by census block groups indicating census block groups above 95th percentile of national values for wastewater discharge
tm_shape(la) +
  tm_polygons(fill = 'P_PWDIS',
              title = "Percentile of Wastewater Discharge") +
  tm_shape(los_angeles_water) +
  tm_dots( 'P_PWDIS',
    alpha = 0.7,
    title.size = "A",
    col = 'red',
    title = 'A') +
  tm_layout(
    legend.outside = FALSE,
    main.title = 'LA Wastewater discharge above 95th percentile',
    main.title.size = 10,
    legend.text.size = 0.7,
    legend.title.size = 0.7
  ) +
  tm_compass(type = "8star",
             size = 1.5,
             position = c("left","bottom")) +
  tm_graticules() + 
  tm_scale_bar(position = c(0.7,0)) +
  tm_add_legend(labels = 'Wastewater Discharge',
                fill = 'red')
  




```

Find the percent of census block groups that have:\
- less than 5% of the population is considered low income (5 points)

```{r }
# Number of census block groups that have less than 5% of the population 
# considered low income
low_income_5 <- los_angeles %>%
  select(LOWINCPCT) %>%
  filter(LOWINCPCT < .05) 


# 6.11% of the population is low income
print(nrow(low_income_5)/nrow(los_angeles) * 100)


```

Find the percent of census block groups that are:\
- above the 80th percentile for Particulate Matter 2.5 AND\
- above the 80th percentile for Superfund proximity (10 points)

```{r }
# above the 80th percentile for Particulate Matter 2.5 and above the 80th percentile for Superfund proximity
pm2_5 <- los_angeles %>% select(P_PM25, P_PNPL) %>%
  filter(P_PM25 > 80 & P_PNPL > 80) 

# 17.357% of the census block are above the 80th percentile for Particulate 
# Matter 2.5 and above the 80th percentile for Superfund proximity
nrow(pm2_5)/nrow(los_angeles)*100
```

Import redlining information for Los Angeles.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>%
  st_make_valid()

```
Make a map of historical redlining boundaries, colored by HOLC grade. (5
points)

```{r }
# Map of historical redlining boundaries colored by HOLC grade

tm_shape(LA_redlining) +
  tm_polygons("holc_grade",
              title = "HOLC grade") +
  tm_compass(type = "8star",
             size = 2) +
  tm_graticules() +
  tm_scale_bar() + 
  tm_layout(main.title = "LA redlining boundaries")

```

Find the number of census block groups that fall within areas with HOLC
grades hint: make sure the CRS match (15 points)

```{r eval=FALSE}
# Check CRS of LA_redlining
st_crs(LA_redlining)

# Check CRS of Los Angeles
st_crs(los_angeles)
```

**-2 did not remove census block duplicates**

```{r eval=TRUE}
# Transform LA_redlining to ESPG:3857
LA_redlining <- st_transform(LA_redlining, "EPSG:3857")

# Find the number of census blocks groups that fall within areas with
# HOLC grades

census_block <- st_intersection(x = los_angeles, y = LA_redlining)

# 6090 census block groups fall within areas with HOLC grades
paste(nrow(census_block),"census block groups")
```

Summarize current conditions based on EJScreen data within historical
redlining categories using the mean of the following variables:\
-% low income.\
- percentile for particulate Matter 2.5.\
- percentile for low life expectancy. 45% - percentile for air toxics
cancer risk (20 points) 52%

```{r }

# Find the mean of % low income within historical redlining categories

mean_table <- census_block %>% 
  group_by(holc_grade) %>% 
  summarise(mean_Low_Income = mean(LOWINCPCT, na.rm = TRUE),
            mean_particular_matter = mean(P_PM25, na.rm = TRUE),
            mean_low_life_expectancy = mean(P_LIFEEXPPCT, na.rm = TRUE),
            mean_cancer_risk = mean(P_CANCER, na.rm = TRUE)
            ) %>% 
  select(-Shape) #remove shape column 

# Table with values
formattable(mean_table)
```

Please write 2-3 sentences reflecting on these results. (5 points)

For all of the variables, they have a higher mean in Holc Grades C and D. Areas that were deemed "Definitely Declining"(C) and "Hazardous"(D) have higher
mean values in % low income, particular Matter 2.5, percentile for low life expectancy,and percentile for air toxic cancer risk. Areas C and D have worst quality of life.

### Investigate the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC
grads, find the percent of observations within each redlining categories
and plot results. hint: make sure that the bird observations have the
same CRS as redlining data. (20 points) \`\`\`\`{r }

\`\`\`

```{r}
# import bird observation data 
bird_observation <- st_read("~/eds223/assignment-2-hazelvaq/data (1)/gbif-birds-LA")
```


```{r eval=FALSE}
# check bird observation CRS
st_crs(bird_observation)

# check redlining CRS
st_crs(LA_redlining)
```


```{r}
# transform bird observation CRS to EPSG 3857
bird_observation <- st_transform(bird_observation,"EPSG:3857")

# Intersect red lining map and bird observation
la_birds <- st_intersection(x = LA_redlining, y = bird_observation)

# Percent observations of birds in HOLC grads
birds <- la_birds %>% 
  filter(year == 2022) %>% 
  group_by(holc_grade) %>% 
  summarise(count = n()) %>% 
  mutate("sum" = sum(count),
         "percent" = count / sum * 100)

## Plot birds observations
ggplot(birds) + 
  geom_col(aes(x = holc_grade,
               y= percent, 
               fill = holc_grade)) +
  labs(x = "HOLC grade",
       y = "Percent of Observations (%)",
       title = "2022 Bird Observations within neighborhoods with HOLC grades",
       fill = "HOLC grade")
```


Please write 2-3 sentences reflecting on these results. Are they
surprising? If so, why? What are some potential implications? (5 points)

HOLC grades C and D had the highest percent of bird observations. This is surprising as these areas are deemed "worst" and have higher means in particular matter 2.5. I would expect there to be more birds in grade A or B instead. Some potential implications could be that there are just more people in grade C and D reporting the birds. Or there could just be more people in general living in Grade C and D then A and B. 


