---
title: 'Analyzing the affect of Redlining on LAs reported bird biodiversity'
author: "{Hazel Vaquero}"
output: html_document
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```


# Background Information

Redlining is a historical practice with racial undertones incorporated
in 1930 by the Home Owners' Loan Corporation (HOLC), as part of the New
Deal. Neighborhoods were rated based on their perceived safety for real
estate investment. Their ranking system, A (green), B (blue), C
(yellow), D (red)) was then used to block access to loans for home
ownership. This practice targeted and discriminated against minority
residents. This unjust practice halted economic development and created
environmental injustice[^1].

A recent study found that redlining is also affecting the observations
of biodiversity[^1]. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most under sampled areas across 195 US cities.
Observation like these rely on community or citizen scientist to
contribute and identify species. The gap in data is concerning because
conservation decisions are made based on these data.

[^1]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Specifically in LA county, neighborhoods in redlining district view less
biodiversity in birds[^2]. These areas have fewer canopy covers which
leads to only certain birds more resilient to urban life to be present.

[^2]: Eric M Wood, Sevan Esaian, Christian Benitez, Philip J Ethington,
    Travis Longcore, Lars Y Pomara, Historical racial redlining and
    contemporary patterns of income inequality negatively affect birds,
    their habitat, and people in Los Angeles, California, Ornithological
    Applications, 2023;, duad044,
    <https://doi.org/10.1093/ornithapp/duad044>

# Data

[EJScreen]{.underline}

The Environmental Protection Agency's Environmental Justice Screening and
Mapping Tool spatial data was used to access data on income, health, and race for 
LA county. The EJScreen uses data that is publicly available to be used to 
analyze environmental justice across the country. There are potential limitations
to this dataset: uncertainty in estimates for smaller areas, does not cover all 
relevant issues, and environmental indicators are only screening-level proxies
for actual health impacts.

EJScreen provides data on environmental and demographic information for
the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract)
and [block group](https://en.wikipedia.org/wiki/Census_block_group)
levels.

[Mapping Inequality]{.underline}

The redlining maps were obtain from the [Digital Scholarship
Lab](https://dsl.richmond.edu/). Researchers at the University of
Richmond have digitized maps and information from the HOLC as part of
the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project. It was directly accessed from the site and downloaded as a
GeoJSn

Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads)[^3].

[^3]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

[Global Biodiversity Information Facility]{.underline}

The community scientist bird observations data was downloaded as a
ShapeFile from the [Global Biodiversity Information Facility](gbif.org).
Observations include the location, date, and species observed..

# Analysis:

My initial analysis is to get a scope of LA county economic and racial
distribution. Using the EJScreen I am interested in analyzing Census
block groups that:

    -   are above the 95th percentile of national values for wastewater
        discharge

    -   have less than 5% of the population are considered low-income

    -   above the 80th percentile for Particular Matters 2.5

    -   above the 80th percentile for Superfund proximity

Building on to these I will also analyze the economic and health
disparity in historical redlined areas. How does redlining areas relate
to:

-   rates of low income

-   percentile for Particular Matter (PM) 2.5

-   percentile for low life expectancy

-   percentile for air toxic cancer risk

## Import Libraries

```{r, message=FALSE, warning=FALSE}
library(tmap)
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)
library(gt)
library(magrittr)
library(formattable)
library(janitor)
```

## Import data

Import our EJScreen, Redlining GeoJSn data, and bird observations.

```{r, message=FALSE}
# Import EJ Screen data
ej_screen <-
  st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb") %>% 
  clean_names() # clean column names

# Import redlining information for Los Angeles

LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>% 
  st_make_valid() # validates geometry

# Import bird observation data
bird_observation <- st_read("data/gbif-birds-LA/gbif-birds-LA.shp")

```

## Prepare data

Let's prepare the data for Los Angeles county and Redlining for
analysis.

```{r, message=FALSE, results='hide'}
# Filter to Los Angeles and remove NA values
los_angeles <- ej_screen %>% filter(cnty_name == "Los Angeles County",
                                    !is.na(p_pwdis))

# Update CRS of Los Angeles and Redlining
if (st_crs(los_angeles) != st_crs(LA_redlining)){
  LA_redlining <- st_transform(LA_redlining, crs = st_crs(los_angeles))
}

# Check if CRS are equal
st_crs(los_angeles) == st_crs(LA_redlining)
```

### EJScreen parameters analysis
The first parameter we are interested in is the census block groups that 
are above the 95th percentile of national values for wastewater discharge. 

```{r, include=FALSE}
# Filter census block groups
los_angeles_water <- los_angeles %>% 
  filter(p_pwdis > 95) %>% 
  st_centroid() # turn into a point polygon
```

Once filtered we can map our results 

```{r}
# Los Angles centroid base
tm_shape(los_angeles) +
  tm_polygons(fill = 'p_pwdis',
              title = "Percentile of Wastewater Discharge") +
  tm_shape(los_angeles_water) + #wastewater discharge centroid
  tm_dots(
    'p_pwdis',
    alpha = 0.7,
    title.size = "A",
    col = 'red',
    title = 'A'
  ) +
  tm_layout(
    legend.outside = FALSE,
    main.title = 'LA Wastewater discharge above 95th percentile',
    main.title.size = 10,
    legend.text.size = 0.7,
    legend.title.size = 0.7
  ) +
  tm_compass(type = "8star",
             size = 1.5,
             position = c("left", "bottom")) +
  tm_graticules() +
  tm_scale_bar(position = c(0.7, 0)) +
  tm_add_legend(labels = 'Wastewater Discharge',
                fill = 'red')
```

The next environmental justice parameters I want to analyze are what percentage 
of census block groups have less than 5% of the population considered low income?

```{r}
# filter census blocks
low_income_5 <- los_angeles %>%
  select(lowincpct) %>%
  filter(lowincpct < .05) 


# Percent of the population is low income
paste(round(nrow(low_income_5)/nrow(los_angeles) * 100,3), "% of the population is low income")
```

As well as, what percentage of census block are above the 80th percentile for 
PM 2.5 and Superfund proximity.

```{r}
# filter PM 2.5 and Superfund proximity
pm2_5 <- los_angeles %>% select(p_pm25, p_pnpl) %>%
  filter(p_pm25 > 80 & p_pnpl > 80) 

# Percent of census blocks
paste(round(nrow(pm2_5)/nrow(los_angeles)*100,3), "% of the census block are 
above the 80th percentile for PM 2.5 and Superfund proximity")
```
### Historical redlining paramter analysis

Next let's make a map of the historical redlining boundaries, colored by HOLC grade 
for LA county.

```{r}
tm_shape(LA_redlining) +
  tm_polygons("holc_grade",
              title = "HOLC grade") +
  tm_compass(type = "8star",
             size = 2) +
  tm_graticules() +
  tm_scale_bar() + 
  tm_layout(main.title = "LA redlining boundaries")
```
First we need to identify the number of sensus block groups that fall within areas
with HOLC grades.

```{r eval=TRUE}
# Filter LA EJScreen for redlining data

redlining_blocks <- los_angeles %>%
  st_filter(y = LA_redlining, .predicate = st_intersects)

# Census blocks
paste(nrow(redlining_blocks),"census block groups")
```

Combining the Redlining and EJScreen data I am interested in analyzing how current 
conditions are within historical redlining categories. I will analyze the mean
for:
-% low income
- percentile for particulate Matter 2.
- percentile for low life expectancy 
- percentile for air toxics cancer risk

```{r }

# Combine LA EJScreen and Redlining data for LA county

LA <- st_intersection(los_angeles, LA_redlining)

# Find the mean of % low income within historical redlining categories
mean_table <- LA %>% 
  group_by(holc_grade) %>% 
  summarise(mean_Low_Income = mean(lowincpct, na.rm = TRUE),
            mean_particular_matter = mean(p_pm25, na.rm = TRUE),
            mean_low_life_expectancy = mean(p_lifeexppct, na.rm = TRUE),
            mean_cancer_risk = mean(p_cancer, na.rm = TRUE)
            ) %>% st_drop_geometry() # drop geometry
  
# Table with values
formattable(mean_table)
```
For all of the variables, they have a higher mean in Holc Grades C and
D. Areas that were deemed "Definitely Declining"(C) and "Hazardous"(D)
have higher mean values in % low income, particular Matter 2.5,
percentile for low life expectancy,and percentile for air toxic cancer
risk. Areas C and D have worst quality of life. 

### Bird observatory

The final analysis this notebook will cover is analyze if areas in historical
redlining report fewer observations of birds?

```{r}
# Transform bird observation CRS to redlining
bird_observation <- st_transform(bird_observation,crs = st_crs(LA_redlining))

# Intersect red lining map and bird observation
la_birds <- st_intersection(x = LA_redlining, y = bird_observation)

# Percent observations of birds in HOLC grads
birds <- la_birds %>% 
  filter(year == 2022) %>% 
  group_by(holc_grade) %>% 
  summarise(count = n()) %>% 
  mutate("sum" = sum(count),
         "percent" = count / sum * 100)

## Plot birds observations
ggplot(birds) + 
  geom_col(aes(x = holc_grade,
               y= percent, 
               fill = holc_grade)) +
  labs(x = "HOLC grade",
       y = "Percent of Observations (%)",
       title = "2022 Bird Observations within neighborhoods with HOLC grades",
       fill = "HOLC grade")

```
HOLC grades C and D had the highest percent of bird observations. This is 
surprising as these areas are deemed "worst" and have higher means in particular
matter 2.5. I would have expected there to be more birds reported in grade A or
B instead. Some potential implications could be that there are just more people 
in grade C and D reporting the birds. Or there could just be more people in 
general living in Grade C and D then A and B. 



